\chapter*{Introduction to This Textbook}

Welcome to Stage 5 Science! This textbook has been specially designed to build on your Stage 4 knowledge and help you develop deeper scientific understanding and skills. Whether you're a student who loves to explore scientific concepts in depth, someone who thinks about ideas in unique ways, or a learner who appreciates clear explanations and engaging activities, this book will support your journey.

\section*{How to Use This Book}

As you explore this textbook, you'll notice several special features:

\begin{itemize}
    \item \textbf{Margin Notes:} Extra information, definitions, and extension ideas in the margins.
    \item \textbf{\challengeicon\ Challenge Questions:} Advanced questions that stretch your thinking beyond the basics.
    \item \textbf{Tiered Activities:} Choose your level of challenge—core questions for everyone and advanced options to push your thinking.
    \item \textbf{Investigations:} Hands-on experiments to develop your scientific inquiry skills.
    \item \textbf{Real-World Scenarios:} Apply your scientific knowledge to authentic situations, similar to those you might encounter in PISA assessments.
    \item \textbf{Extension Activities:} Optional explorations for students ready to go beyond the standard curriculum.
\end{itemize}

This textbook aligns with the NSW Stage 5 Science syllabus while providing enrichment opportunities and support for diverse learning styles. It also helps prepare you for PISA assessments by developing your ability to apply scientific knowledge to real-world situations.

Science at Stage 5 builds on your previous learning and prepares you for future studies or careers in science-related fields. As you engage with the concepts and activities in this book, remember that science is both a body of knowledge and a process of inquiry—continue to observe, question, investigate, and explain the fascinating world around you!